visualFields 1.0.7: Tools for Analyzing Visual Fields in \proglang{R}

### 2025-06-16 jss-article-rnw/article.Rnw
Added captions and did some copyediting.

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}

@@

\title{visualFields 1.0.7: Tools for Analyzing the Field of Vision in \proglang{R}}
\Plaintitle{visualFields 1.0.7: Tools for Analyzing the Field of Vision in R}
\Shorttitle{visualFields 1.0.7 in \proglang{R}}

\Abstract{
  The \pkg{visualFields} package is part of the Open Perimetry Initiative (OPI), 
  a standard for interfacing with visual field testing machines. This article 
  introduces the package, which provides tools for analyzing the field of vision, 
  including visualization, statistical analysis, and clinical interpretation of 
  visual-field loss and its change over time.
}

@@

The Open Perimetry Initiative (OPI) \citep{IMF:Tur2012, IMF:Marin2022a} is an open standard for interfacing with visual field testing devices (perimeters). It defines a set of core functions that enable the construction of a wide range of visual field tests. As of October 2017, OPI is fully implemented on the Octopus 900 and partially supported on the Heidelberg Edge Perimeter, Kowa AP 7000, CrewT IMO, and Centervue Compass.

@@

The visual field refers to the entire area that can be seen at a given moment, encompassing both central and peripheral vision. It is typically measured in degrees of visual angle, with the fixation point defined as the origin at $x = 0^\circ$ and $y = 0^\circ$. Locations temporal and superior to fixation are represented by positive angles, while those nasal and inferior are denoted by negative angles.

In healthy vision, most visual functions, particularly light sensitivity, decline with increasing eccentricity—i.e., distance from the center of fixation. This decline in differential light sensitivity forms a three-dimensional structure known as the hill of vision, also referred to as Traquair’s island of vision \citep{IMF:Traquair1938}. Figure~\ref{fig:hill} below shows a classic rendering of the hill of vision by Aulhorn and Harms \citep{IMF:Aul1967}.

\begin{figure}[t!]
\centering
\includegraphics[width=0.75\textwidth]{AulhornHillOfVision.png}
\caption{\label{fig:hill} Hill of vision as rendered by Aulhorn \& Harms.}
\end{figure}

@@

In visual field testing, the observer either presses or does not press a button within a fixed time window following the presentation of a light increment in the visual field (a yes/no response). The relationship between stimulus strength and response probability is modeled by a \emph{psychometric function}, which characterizes the likelihood of a "yes" response as a function of stimulus strength \citep{CH:knoblauch_modeling_2012}. These models yield an estimate of the threshold, defined as the stimulus strength at which the observer detects the stimulus with a probability of 50\%.

@@

By default, the package is set up to handle Humphrey Field Analyzer 24-2 visual fields. It is recommended to use such data for exploring the package and data is provided by the package. However, other patterns and devices can be used. The package contains all necessary data for the HFA 10-2 pattern. The package is designed to be extensible, allowing users to add new patterns, devices, and normative datasets.

@@

\section{Methods} \label{sec:methods}

@@

The package includes example data from a glaucoma patient (\code{vfpwgSunyiu24d2}), consisting of 42 visual field tests.

The first ten columns contain metadata for each test, including reliability indices (false negatives, false positives, and fixation losses). The remaining columns contain the actual test values (sensitivities) at the locations defined in the \code{locmap}.

<<>>=
str(vfpwgSunyiu24d2[, 1:10], strict.width = "wrap")
@

The \code{date} variable is stored as a \code{POSIXct} object, which can be converted to a date format using the \code{as.Date()} function. Each row represents a monocular test.

The eye is labeled using Latin abbreviations:

\begin{description}
\item [OD] oculus dexter -> right eye
\item [OS] oculus sinister -> left eye
\item [OU] oculus uterque -> both eyes
\end{description}

The \code{type} variable allows assignment of tests to different groups.

@@

Deviations from the expected psychometric function may arise due to observer inattention or short-term threshold fluctuations. In diseases like glaucoma, short-term variability is often elevated in damaged regions of the visual field. Response quality is typically assessed by:

\begin{description}
  \item [False negatives (FNR)] Missed responses to stimuli that were previously seen.
  \item [False positives (FPR)] Responses when no stimulus is presented
  \item [Fixation losses] Responses to stimuli presented within the blind spot
\end{description}

These metrics help assess the reliability of a test.

\subsubsection{Psychophysical data: sensitivities}

Initial columns store patient information and reliability metrics. Actual test values (sensitivities) begin at the column number stored in the \code{locini} variable. The function \code{getvfcols()} returns the numbers of the columns that contain the actual test values at the locations \code{l1} to \code{ln}, where 1..n corresponds to the rows of \code{coords} in the \code{locmaps}.

@@

The \code{vfread()} and \code{vfread()} functions can be used to read and write data in this format from a CSV file. However, these functions cannot be used to import native data from the perimeters. For this, the package includes functions to import data from the Octopus 900 and the Humphrey Field Analyzer.

\begin{itemize}
  \item \code{loadoctopus()} for Octopus 900 (Haag-Streit, Switzerland)
  \item \code{loadhfaxml()}, \code{loadhfadicom()}, \code{loadhfaxmlbatch()}, \code{loadhfadicombatch()} for the Humphrey Field Analyzer (Zeiss)
\end{itemize}

Detailed examples for exporting data from the Octopus 900 and using \code{loadoctopus()} can be found here: \url{https://huchzi.github.io/vf1/loadOctopus.html}.

@@

<<>>=
names(locmaps)
@

@@

Other patterns are available in the package, including the \code{10-2} grid.

<<>>=
ten_two_locmap <- locmaps$p10d2
str(ten_two_locmap, strict.width = "wrap")
@

@@

<<>>=
str(
  vfctrSunyiu24d2, 
  strict.width = "wrap")
@

@@

<<>>=
str(
  normvals$sunyiu_24d2, 
  max.level = 1, strict.width = "wrap")
@

To build the normative model for a custom dataset, use \code{nvgenerate()}.

@@

<<>>=
head(nv$agem$coeff, 5)
@

@@

\section{Results} \label{sec:results}

@@

<<>>=
gl <- getgl(example2)
str(gl, max.level = 1, strict.width = "wrap")
gl[, -(1:getlocini() - 1)]
@

@@

<<>>=
getglp(gl)[, -(1:getlocini() - 1)]
@

@@

There are various ways to visualize perimetric data. These visualizations should be interpreted as simplified representations of the hill of vision.

Standard displays include color-coded probability maps, graphical representations of the blind spot, and border enhancements that encode diagnostic information. For example, \code{vfplot} with the argument 	\code{"td"} generates plots that require normative data on differential light sensitivity and the probabilities of deviations from these norms. This information is included for the Humphrey Field Analyzer patterns but must be provided separately for other perimeters.

By convention, when displaying both eyes side-by-side, the left eye appears on the left and the right eye on the right. This reflects the patient's visual fields, with blind spots positioned temporally. This is in contrast to structural imaging, which is typically displayed as viewed by the examiner.

It is important to specify the type of data represented in the visual field tables. The default is set to \emph{total deviation} (	exttt{td}).

@@

In this plot, yellow indicates a probability $<$ 0.05, orange $<$ 0.02, red $<$ 0.01, and dark red $<$ 0.005.

@@

Monitoring visual function over time is a key application of perimetry in managing glaucoma.

Sensitivity measurements naturally fluctuate between tests. This variability increases in eyes with glaucoma, especially at affected locations. Hence, two tests showing differences cannot alone indicate progression. Instead, statistical analysis across a series of tests is required to detect significant change.

Progression analysis methods fall into two broad categories: \emph{event-based} and \emph{trend-based} approaches. These may further be classified by their reliance on population-based normative data, mathematical models of progression, or model-free methods.

@@

The \code{glr()} function estimates a regression model for global indices. For the example series, the patient loses \code{getgl(example3)} dB/year, with $p < getgl(example3)$.

%\code{r round(abs(glr(getgl(example3))$sl), 2)}
%\code{r format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)}

@@

To visualize these results:

In this display, yellow indicates $p < 0.05$, orange $< 0.02$, red $< 0.01$, and dark red $< 0.005$.

@@

@@

A plot of the 10-2 grid and corresponding ganglion cell bodies:

@@

plot_grid <- function() {
  lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
  lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
  lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
  lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
  lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
  lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
  for(r in c(2.5, 5, 7.5, 10))
    draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
  text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
  text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
}

par(mfrow = c(1, 2))
plot(x = locmap$coord$x, 
     y = locmap$coord$y, 
     type = "n", 
     asp = 1,
     xlim = c(-10, 12),
     ylim = c(-10, 10),
     yaxt = 'n',
     xaxt = 'n',
     bty = 'n',
     xlab = "",
     ylab = "",
     main = "10-2 grid of test locations")
plot_grid()
points(x = locmap$coord$x, y = locmap$coord$y, bg = "white")

plot(x = locmap$coord$x, 
     y = locmap$coord$y, 
     type = "n", 
     asp = 1,
     xlim = c(-10, 12),
     ylim = c(-10, 10),
     yaxt = 'n',
     xaxt = 'n',
     bty = 'n',
     xlab = "",
     ylab = "",
     main = "Corresponding ganglion cells")
plot_grid()
points(x = gcmap$x, y = gcmap$y, pch = 21, bg = "white")
@

Nerve fiber paths from each location to the optic disc can also be modeled, using Jansonius' framework \citep{IMF:Jan2009, IMF:Jan2012}.

<<>>=
psi23 <- loc2psi(vf2gc(locmaps$p24d2$coord[23,]))
ang23 <- psi2oct(psi23)

ang23 - 360
@

\subsection{DICOM support}

DICOM standards for static perimetry are defined in Supplement 146. The package includes preliminary functions for reading HFA-generated DICOM files, but exporting to DICOM is not yet supported.

\subsection{Support for other perimeters}

Perimetry conventions vary across manufacturers. The package includes normative data for Humphrey 24-2 (Goldmann III) and Iowa full-field perimetry with Goldmann V \citep{IMF:Marin2018c}. The \code{loadoctopus} function provides partial support for Octopus G1 fields, but normative support is incomplete.

Users may provide custom normative datasets using \code{nvgenerate()} and \code{setnv()}.

\subsection{Setting the environment}

Each analysis requires environment variables:
\begin{itemize}
\item \code{nv}: normative values
\item \code{locmap}: test grid
\item \code{gpar}: graphical parameters
\item \code{locini}: first sensitivity column
\end{itemize}

<<>>=
names(locmaps)
names(normvals)
names(gpars)
@

These must be consistent. For example:

<<>>=
setlocmap(locmaps$pPC26)
setnv(normvals$iowa_PC26_pw)
setgpar(gpars$pPC26)
setlocini(24)
@

To reset to default (Humphrey 24-2):

@@

\subsection{Custom configurations}

Custom test structures must:
\begin{itemize}
\item Start with columns: \code{id}, \code{eye}, \code{date}, \code{time}
\item Include \code{age}, optionally \code{fpr}, \code{fnr}, \code{fl}
\item Have sensitivity columns named \code{l1}--\code{lk}
\end{itemize}

To import custom locations:

<<eval=FALSE>>=
g1 <- list(coord = read.csv("./g1.csv"), 
           name = "G1", 
           desc = "G1 map", 
           bs = numeric())
setlocmap(g1)
setgpar(vfgpar(g1$coord))
vfplot(vfread("./vfg1.csv"), type = "s")
@

If normative data exist, use \code{vfread()}, \code{nvgenerate()}, and \code{setnv()} to enable deviation computations.

@@

\newpage

\begin{appendix}

\section{Clinical aspects} \label{app:clinical}

\subsection{How is perimetry performed?}

In a typical SAP test, the patient is seated comfortably in front of a perimeter and places their head on a chin rest while looking into a white Ganzfeld dome. The room is dimly lit and quiet, and the patient is briefed by an experienced technician before the examination. During the test, the patient fixates on a central target (e.g., a dot, cross, or diamond), and presses a button whenever a light stimulus is perceived.

Stimuli are presented briefly (typically 100–200 ms), minimizing the opportunity for eye movement. A response is only considered valid if it occurs within a defined time window following the stimulus. A computer presents stimuli at discrete locations in a grid, in randomized order. The stimulus intensity at each location is adjusted adaptively based on the patient' responses. Because of randomization, the patient cannot anticipate where the next stimulus will appear.

The figure below shows the commonly used 24-2 grid of test locations. The gray ellipse represents the anatomical blind spot; the red points fall within this region and are expected to show no sensitivity.

The grid, stimulus size, presentation duration, and intensity algorithm remain constant throughout the test and together define the perimetry program. It is important to note that results from different perimetry programs are not directly comparable.

\subsection{Clinical applications of perimetry}

In clinical settings, perimetry is used to detect and monitor visual field defects or scotomas, which can arise from damage anywhere along the visual pathway—from the ocular media (cornea, lens, vitreous), to the retina, optic nerve, or visual cortex.

The spatial pattern of visual field loss often provides diagnostic clues. For instance, a right occipital lobe infarct can result in left homonymous hemianopia—complete loss of the left visual field in both eyes, with a sharp border along the vertical meridian. Conversely, glaucoma typically presents as arcuate defects, often beginning in the upper or lower hemifield, reflecting characteristic damage to the optic nerve head.

Perimetry supports not only diagnosis but also evaluation of functional impact (e.g., fitness to drive) and disease progression monitoring.

\section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}

\begin{leftbar}
References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}). All
references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|,
\verb|\citealp| etc.\ (and never hard-coded). This commands yield different
formats of author-year citations and allow to include additional details (e.g.,
pages, chapters, \dots) in brackets. In case you are not familiar with these
commands see the JSS style FAQ for details.

Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task -- especially
when acquiring the entries automatically from mixed online sources. However,
it is important that informations are complete and presented in a consistent
style to avoid confusions. JSS requires the following format.
\begin{itemize}
  \item JSS-specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
    be used in the references.
  \item Titles should be in title case.
  \item Journal titles should not be abbreviated and in title case.
  \item DOIs should be included where available.
  \item Software should be properly cited as well. For \proglang{R} packages
    \code{citation("pkgname")} typically provides a good starting point.
\end{itemize}
\end{leftbar}

\end{appendix}

>>>>

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE, width=7, height=5, out.width='0.1\\textwidth'}

@@

\title{visualFields 1.0.7: Tools for Analyzing Visual Fields in \proglang{R}}
\Plaintitle{visualFields 1.0.7: Tools for Analyzing Visual Fields in R}
\Shorttitle{visualFields 1.0.7 in \proglang{R}}

\Abstract{
  The \pkg{visualFields} package is part of the Open Perimetry Initiative (OPI), 
  a standard for interfacing with visual field testing machines. This article 
  introduces the package, which provides tools for analyzing the field of vision, 
  including visualization, statistical analysis, and clinical interpretation of 
  visual-field loss and its progression over time.
}

@@

The Open Perimetry Initiative (OPI) \citep{IMF:Tur2012, IMF:Marin2022a} provides an open standard for interfacing with visual field testing devices (perimeters). It defines a set of core functions that enable the construction of a wide range of visual field tests. As of October 2017, OPI has been fully implemented on the Octopus 900 and partially supports the Heidelberg Edge Perimeter, Kowa AP 7000, CrewT IMO, and Centervue Compass.

@@

The visual field refers to the entire area that can be seen at a given moment, encompassing both central and peripheral vision. It is typically measured in degrees of visual angle, with the fixation point defined as the origin at $x = 0^\circ$ and $y = 0^\circ$. Locations temporal and superior to fixation are assigned positive angles, whereas nasal and inferior locations are assigned negative angles.

In healthy vision, most visual functions, particularly light sensitivity, decline with increasing eccentricity—i.e., distance from the center of fixation. This decline in differential light sensitivity gives rise to a three-dimensional structure known as the hill of vision, also referred to as Traquair’s island of vision \citep{IMF:Traquair1938}. Figure~\ref{fig:hill} below shows a classic rendering of the hill of vision by Aulhorn and Harms \citep{IMF:Aul1967}.

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{Figure1.pdf}
\caption{\label{fig:hill} Perimetry. A) A (manual) perimeter with an inset showing the stimulus enlarged. B) The hill of vision.}
\end{figure}

@@

In visual field testing, the observer responds by either pressing or not pressing a button within a fixed time window following the presentation of a light increment in the visual field (a yes/no response). The relationship between stimulus strength and response probability is modeled by a \emph{psychometric function}, which characterizes the likelihood of a "yes" response as a function of stimulus strength \citep{CH:knoblauch_modeling_2012}. These models yield an estimate of the threshold, defined as the stimulus strength at which the observer detects the stimulus with a probability of 50\%.

@@

By default, the package is set up to handle Humphrey Field Analyzer 24-2 visual fields. It is recommended to use such data for exploring the package and data is provided by the package. However, other patterns and devices can be used. The package also contains all necessary data for the HFA 10-2 pattern. However, the package is designed to be extensible, allowing users to add new patterns, devices, and normative datasets.

@@

\section{Humphrey Field Analyser 24-2 pattern} \label{sec:methods}

In this section, we will show how to install and load the package, and how the data supplied by the package can be analyzed with the default environment settings. In the following sections, we will show how to plot the data, perform longitudinal and cross-sectional analyses, how to import data from perimeters, and how to set up the environmnent for these data.

@@

The most important data are the perimetry results. However, the environment needs to be set up correctly for the corresponding type of perimetry. Here, we start with data from a Humphrey Field Analyser (HFA) with the 24-2 pattern that is supplied by the package (\code{data(vfpwgSunyiu24d2)}), and demonstrate how the data is structured and what environment settings are used. The default environment settings are aimed at the HFA 24-2, and they can always be reset with \code{setdefaults()}. The data is from a glaucoma patient (\code{vfpwgSunyiu24d2}) and consists of 42 consecutive visual field tests.

<<>>=
data(vfpwgSunyiu24d2)
str(vfpwgSunyiu24d2[, 1:10], strict.width = "wrap")
@

The first ten columns contain metadata for each test, including reliability indices. The remaining columns contain the actual test values (sensitivities) at the locations defined in the \code{locmap}. The \code{date} variable is stored as a \code{POSIXct} object, which can be converted to a date format using the \code{as.Date()} function. Each row represents a monocular test. The \code{type} variable allows assignment of tests to different groups. The eye is labeled using Latin abbreviations:

\begin{description}
\item [OD: oculus dexter] right eye
\item [OS: oculus sinister] left eye
\item [OU: oculus uterque] both eyes
\end{description}

@@

Deviations from the expected psychometric function may arise due to observer inattention or short-term threshold fluctuations. In diseases like glaucoma, short-term variability is often elevated in damaged regions of the visual field. Response quality is typically assessed by following metrics and these help to assess the reliability of a test. :

\begin{description}
  \item [False negatives (FNR)] Missed responses to stimuli that were previously seen.
  \item [False positives (FPR)] Responses when no stimulus is presented when the patients expects a presentation from the rhythm of the exam
  \item [Fixation losses] Responses to stimuli presented within the blind spot
\end{description}

\subsubsection{Local sensitivities}

Actual test values (sensitivities) begin at the column number stored in the \code{locini} variable of the environment. The function \code{getvfcols()} returns the numbers of the columns that contain the actual test values at the locations $l_{1}$ to $l_{n}$, where n corresponds to the number of locations tests (n = 54 for the 24-2 pattern).

@@

Here, the visual field data is not value, because an illegal column was added. The \code{vfread()} and \code{vfread()} functions can be used to read and write data in this format from a CSV file, but they cannot be used to import native data from the perimeters.

The following functions can be used to import data from perimeters.

\begin{itemize}
  \item \code{loadhfaxml()}, \code{loadhfadicom()}, \code{loadhfaxmlbatch()}, \code{loadhfadicombatch()} for the Humphrey Field Analyzer (Carl Zeiss Meditec, Dublin, CA, USA)
  \item \code{loadoctopus()} for Octopus 900 (Haag-Streit, Köniz, Switzerland)
\end{itemize}

@@

<<>>=
data(locmpas)
names(locmaps)
@

@@

@@

<<>>=
data(fvctrSunyiu24d2)
str(
  vfctrSunyiu24d2[, 1:getlocini()+5], 
  strict.width = "wrap")
@

@@

<<>>=
data(normvals)
str(
  normvals$sunyiu_24d2, 
  max.level = 1, strict.width = "wrap")
@

@@

<<>>=
nv <- getnv()
head(nv$agem$coeff, 5)
@

@@

\section{Analyzing visual fields} \label{sec:results}

After you have familiarized yourself with the structure of the default data, you will now learn how to use the visual field package to analyze single locations, how to calculate global parameters that summarize the visual field, how plot fields, how to perform longitudinal analyses, and how to do structure-function correlations. Then, in the following section, we will show how to use the package with different patterns and data from different perimeters.

@@

<<>>=
gl <- getgl(example2)
str(gl, max.level = 1, strict.width = "wrap")
gl[, getlocini():ncol(gl)]
@

@@

<<>>=
getglp(gl)[, getlocini():ncol(gl)]
@

@@

There are various ways to visualize perimetric data. These visualizations should be interpreted as simplified representations of the hill of vision. Standard displays include color-coded probability maps, graphical representations of the blind spot, and border enhancements that encode diagnostic information. For example, \code{vfplot} with the argument \code{"td"} generates plots that require normative data on differential light sensitivity and the probabilities of deviations from these norms. This information is included for the Humphrey Field Analyzer patterns but must be provided separately for other perimeters.
By convention, when displaying both eyes side-by-side, the left eye appears on the left and the right eye on the right. This reflects the patient's visual fields, with blind spots positioned temporally. This is in contrast to structural imaging, which is typically displayed as viewed by the examiner. It is important to specify the type of data represented in the visual field tables. The default is set to \emph{total deviation} (\code{td}). In the following plot, yellow indicates a probability $<$ 0.05, orange $<$ 0.02, red $<$ 0.01, and dark red $<$ 0.005.

<<fig=TRUE, out.width='.25\\textwidth', out.height='!'>>=

@@

@@

Monitoring visual function over time is a key application of perimetry in managing glaucoma. Sensitivity measurements naturally fluctuate between tests. This variability increases in eyes with glaucoma, especially at affected locations. Hence, two tests showing differences cannot alone indicate progression. Instead, statistical analysis across a series of tests is required to detect significant change. Progression analysis methods fall into two broad categories: \emph{event-based} and \emph{trend-based} approaches. These may further be classified by their reliance on population-based normative data, mathematical models of progression, or model-free methods.

@@

The \code{glr()} function estimates a regression model for global indices. For the example series, the the regression can be calculated as follows:

<<>>=
global_indices <- getgl(example3)
global_linear_regression <- glr(global_indices[,c("date", "tmd")])
str(global_linear_regression)
global_linear_regression$sl
global_linear_regression$pval
@

@@

@@

Use \code{vfplotplr()} to visualize results from the pointwise linear regression (PLR). In this figure, yellow indicates $p < 0.05$, orange $< 0.02$, red $< 0.01$, and dark red $< 0.005$.

@@

A plot of the 10-2 grid and the displacement of the corresponding ganglion cell bodies:

@@

rbind(data.frame(label = "10-2 grid", locmap$coord), 
      data.frame(label = "Corresponding ganglion cells", gcmap)) |>
        ggplot(aes(x = x, y = y)) +
        geom_point() +
        facet_wrap(~ label) +
  theme_minimal(12) +
  scale_x_continuous("", limits = c(-10, 10)) +
  scale_y_continuous("", limits = c(-10, 10)) +
  guides(x = "none", y = "none")

Nerve fiber paths from each location to the optic disc can also be modeled, using Jansonius' framework \citep{IMF:Jan2009, IMF:Jan2012} with the functions \code{cart2jpolar}, \code{jpolar2cart}, \code{bundlePath}, \code{loc2psi}, and \code{psi2oct}.

\section{Support for other patterns and perimeters}

The best supported alternativ to the 24-2 pattern is the HFA 10-2 pattern. Remember that you must changes other settings along with the \code{locmap}, including the \code{normvals} and the \code{gpar} parameters. Otherwise, the \code{vfplot()} function will throw an error. It is very important that you set up the environment completely (\code{locmap}, \code{nv}, \code{gpar}).

setlocmap(locmaps$p10d2)
setnv(normvals$sunyiu_10d2)
setgpar(gpars$p10d2)
vfplot(vffilter(vfctrSunyiu10d2, id == 2, eye == "OD"))

Remember that you can always reset the environment with...

@@

\subsubsection{Generate normal values}

The locmap will be available online or from the manufacturer. Perimetry conventions vary across manufacturers. The package includes normative data for Humphrey 24-2 (Goldmann III) and Iowa full-field perimetry with Goldmann V \citep{IMF:Marin2018c}. Use \code{nvgenerate()} to build the normative model for a custom dataset. You will need a sufficiently large dataset of normal fields balanced for age in order to create your own norm values.

\subsubsection{Support for Octopus perimeters}

Octopus data can be imported with \code{loadoctopus()}. However, no normal values are provided by the package, yet. Detailed examples for exporting data from the Octopus 900 and using \code{loadoctopus()} can be found here: \url{https://huchzi.github.io/vf1/loadOctopus.html}.

@@
